# Update Data

```{r, echo=FALSE, warning=FALSE, results='hide'}
library(odbc)
library(DBI)
conn <- dbConnect(
  drv = odbc(),
  dsn = "Local Teaching MySQL84",
  database = "ap"
)

# "https://yihui.org/knitr/options/"
library(knitr)
opts_chunk$set(connection = conn)
# row.print works for R chucks, max.print works for SQL chucks
opts_chunk$set(max.print = 15)
```

## Reference

Default database in this chapter is "ap".

Other databases included in the simulation MySQL server:

-   om
-   ex

## Table Operations

### Create table

```{sql, echo=FALSE}
DROP TABLE IF EXISTS invoices_copy
```

```{sql}
CREATE TABLE invoices_copy AS 
-- everything below can be just a standalong query, you can choose to create the new table with only certain data from the original 
SELECT *
FROM invoices
```

Locate the newly created table: invoices_copy in ap database:

```{sql}
SHOW TABLES
FROM ap
```
### Delete Table
Always use full 3-part naming to be sure deleting the correct table
```{sql}
DROP TABLE ap.invoices_copy
```
Confirm deleted
```{sql}
SHOW TABLES
FROM ap
WHERE Tables_in_ap = 'invoices_copy'
```

### Recreate a Table
If you want to remove a table as long as the table exists in the database (Especially useful when you only want to create a temporary table): 
```{sql}
DROP TABLE IF EXISTS invoices_copy
```

## Data Operations
Create backup table for demonstration.
```{sql}
CREATE TABLE invoices_copy AS 
SELECT *
FROM invoices
```
See the schema for the new table. Easier for the adding new data later...
```{sql}
SELECT *
FROM invoices_copy
LIMIT 1
```
Create other backup tables:
```{sql}
DROP TABLE IF EXISTS invoice_archive_copy
```

```{sql}
CREATE TABLE invoice_archive_copy AS 
SELECT *
FROM invoice_archive
```

### INSERT: Add New Records
Add 1 row to invoices_copy table. 
Notice that all column values must be entered, even if it is NULL. 
```{sql}
INSERT INTO invoices_copy VALUES
(115, 97, '456789', '2022-08-01', 8344.50, 0, 0, 1, '2022-08-31', NULL)
```
Check the new record is in place:
```{sql}
SELECT *
FROM invoices_copy 
WHERE invoice_id = 115
```
Add multiple records in one statement:
```{sql}
INSERT INTO invoices_copy VALUES
    (116, 97, '456701', '2022-08-02', 270.50, 0, 0, 1,
    '2022-09-01', NULL),
    (117, 97, '456791', '2022-08-03', 4390.00, 0, 0, 1,
    '2022-09-02', NULL),
    (118, 97, '456792', '2022-08-03', 565.60, 0, 0, 1,
    '2022-09-02', NULL)
```
Verify the new records have been added: 
```{sql}
SELECT *
FROM ap.invoices_copy 
WHERE invoice_id > 115
```

You can also copy and add a whole subset of records from one table (source table) to another (target table). To do this, you need to ensure:

- Target table has all the columns in EXACTLY THE SAME names
- Select only the same columns from the source table  

Example: To copy a whole chunk of rows to invoice_arhchive. 

To start with, the target table invoice_archive_copy has no records...
```{sql}
SELECT * 
FROM invoice_archive_copy
```
Now, we add all invoices (source table) with 0 balance to the archive table.
```{sql}
INSERT INTO invoice_archive_copy
    (invoice_id, vendor_id, invoice_number,
    invoice_total, credit_total, payment_total,
    terms_id, invoice_date, invoice_due_date)
SELECT
    invoice_id, vendor_id, invoice_number,
    invoice_total, credit_total, payment_total,
    terms_id, invoice_date, invoice_due_date
FROM invoices_copy
WHERE invoice_total - payment_total - credit_total = 0
```
Now, let's see the target table again...
```{sql}
SELECT * 
FROM invoice_archive_copy
LIMIT 10
```
### UPDATE: Modify Records

Example: We want to manually update an invoice record with a particular invoice_number

Now, let's see what it looks like before any updates

```{sql}
SELECT *
FROM invoices_copy
WHERE invoice_number = '97/522'
```
Now, we will update the payment_date and payment_total values of the same invoice:
```{sql}
UPDATE invoices_copy
SET payment_date = '2022-09-21', 
    payment_total = 19351.18
WHERE invoice_number = '97/522'
```
Now, let's see what it looks like after the updates:
```{sql}
SELECT *
FROM invoices_copy
WHERE invoice_number = '97/522'
```
We can also update the new values with the calculation from the old values.

Example: We want to increase the credit total by 100
```{sql}
UPDATE invoices_copy
SET credit_total = credit_total + 100 
/* mathmatically, this does not make sense, but here, it simply means credit_total (new) = credit_total (old) + 100 */
WHERE invoice_number = '97/522'
```

Now, let's see what it looks like after the second update:
```{sql}
SELECT *
FROM invoices_copy
WHERE invoice_number = '97/522'
```
You can also use update a batch of records meeting criteria from a subquery.

Example: we want to set all terms_id to 1 (instead of current 2) for a vendor with name called 'Pacific Bell'. This is only possible when querying with vendor table.

Now, let's see what all invoices look with vendor Pacific Bell like, before any updates:
```{sql}
SELECT *
FROM invoices_copy
WHERE vendor_id IN
      (SELECT vendor_id
       FROM vendors
       WHERE vendor_name = 'Pacific Bell')
```
Now, we apply the updates:
```{sql}
UPDATE invoices_copy
SET terms_id = 1
WHERE vendor_id =
      (SELECT vendor_id
       FROM vendors
       WHERE vendor_name = 'Pacific Bell')
```

Now, let's see what all invoices look with vendor Pacific Bell like, after updates:
```{sql}
SELECT *
FROM invoices_copy
WHERE vendor_id IN
      (SELECT vendor_id
       FROM vendors
       WHERE vendor_name = 'Pacific Bell')
```
**NOTE: it is a good idea to query and test the criteria (WHERE clause), before you apply any updates. The same goes to next section "DELETE", when removing records.**


### DELETE: Remove Records
Example: we want to remove all the last 3 inserted records: vendor_id: 116, 117, 118
```{sql}
DELETE FROM invoices_copy
WHERE invoice_id >= 116
```
Now, let's see the result: (all records from 116 and above are gone)
```{sql}
SELECT *
FROM invoices_copy
WHERE invoice_id >= 115
```

```{r, echo=FALSE}
dbDisconnect(conn)
```

