# Working with Numeric Data

```{r, echo=FALSE, warning=FALSE, results='hide'}
library(odbc)
library(DBI)
conn <- dbConnect(
  drv = odbc(),
  dsn = "Local Teaching MySQL84",
  database = "ex"
)

# "https://yihui.org/knitr/options/"
library(knitr)
opts_chunk$set(connection = conn)
# row.print works for R chucks, max.print works for SQL chucks
opts_chunk$set(max.print = 15)
```

## Reference

Default database in this chapter is "ex".

Other databases included in the simulation MySQL server:

-   om
-   ap

## Numeric Data

### Main Types
Integer(INT type): no decimal places.

Fixed-point(DECIMAL type): up to 9 decimal places.

- Because there are finite number of decimal places, they are exact numeric type.

Floating-point (shortly floating, DOUBLE/FLOAT type) values are numeric values with unlimited number of decimal places.

- Because there are unlimited number of decimal places, they are approximate numeric type.

### Other Concepts
Precision: number of digit (e.g. 120.98 has a precision of 5, as in 5 digits in total)

Scale: number of decimal places (e.g. 120.98 has a scale of 2, as in 2 decimal places)

## Numeric Data Manipulation

### Managing Rounding and Scale

```{sql}
SELECT 
  float_value,
  ROUND(float_value, 3),
  TRUNCATE(float_value, 3),
  CEILING(float_value),
  FLOOR(float_value)
FROM float_sample
```
**Note:** negative values are evaluated together with their negativity. 

```{sql}
SELECT 
  CEILING(-12.4567),
  FLOOR(-12.4567)
```
### Calcuation
```{sql}
SELECT 
  float_value,
  ABS(float_value),
  -- SIGN(float_value),
  SQRT(float_value),
  POWER(float_value, 3)
FROM float_sample
```
### Random Number Generation
RAND() function doesn't require any input to work, i.e. they will generate random numbers anyways.

The integer as an argument is called "seed". This is used to ensure everytime the query run it will create exactly the same set of random numbers.

```{sql}
SELECT
  float_value,
  RAND(123)
FROM float_sample
```


```{r, echo=FALSE}
dbDisconnect(conn)
```

